Something something values disclosure replication

### 2022-07-08 paper/paper.Qmd
reproducibility; include emad data in repo
fill in references; working on conclusion

---
title: "Something something values disclosure replication"
author: "Daniel J. Hicks and Emilio Lobato"
abstract: |
  abstract
toc: true
execute:
    echo: false
---

@@

Over the past 15 years, many philosophers of science have rejected the ideal of value-free science and related, traditional understandings of objectivity and political neutrality of science. *[Douglas and Elliott]*  According to the value-free ideal, social and political values — such as feminism, environmentalism, or the protection of human health — have no legitimate role to play in the evaluation of scientific hypotheses.  The value-free ideal is compatible with allowing social and political values to play important roles earlier and later in inquiry.  Specifically, these values may legitimately shape the content and framing of research questions — researchers might decide to investigate whether chemical X causes cancer out of a concern to protect human health — and will be essential when scientific findings are used to inform public policy — say, banning the use of chemical X.  But, according to the value-free ideal, these values must not influence the collection and analysis of data and the process of reaching an overall conclusion about whether or not chemical X causes cancer.

Challenges to the value-free ideal argue that at least some social and political values may, or even should, play a role in the evaluation of scientific hypotheses.  Keeping with the example, the question of whether chemical X causes cancer has much more significant social and political implications than the question of whether chemical X fluoresces under ultraviolet light.  So, in the interest of protecting human health, it would be appropriate to *scientifically* accept the hypothesis that chemical X is carcinogenic based on relatively provisional evidence *[cite]*.

While this kind of argument indicates that at least some social and political values may or should play at least some role in the evaluation of at least some scientific hypotheses, it doesn't provide us with much positive guidance:  which values, playing which roles, in the evaluation of which hypotheses?  *[New Direction]*  One (partial) answer to this set of questions appeals to the ideal of transparency:  scientists should disclose to the public the social and political values that have influenced their research *[Elliott]*.  But some philosophers have objected to this transparency proposal, arguing that it might actually undermine trust in science.  According to this objection, members of the public generally accept the ideal of value-free science; values disclosures would violate this ideal, making scientists (incorrectly) appear biased and untrustworthy *[John, Kovaka]*.

@@

| No other information about the (fictional) Dr. Spence is provided.  The content of the slide varies across the conditions.  Each slide has the header "My conclusion," followed by a list of 2 or 3 bullets.  The first bullet, if present, makes a values disclosure, stating that either economic growth or public health should be a top national priority.  This first bullet is not present in the "no disclosure" values condition.  The second bullet, identical across conditions, states "I examined the scientific evidence on potential health risks of BPA."  The third bullet concludes that BPA either does or does not cause harm to people, depending on whether the subject is in the harm or *[not-harm]* condition.

After viewing the slide, the participant is then asked to rate Dr. Spence's trustworthiness using a semantic differential scale.  @ElliottValuesEnvironmentalResearch2017 did not cite a source for their scale; we used the Muenster Epistemic Trustworthiness Inventory [METI\; @ElliottValuesEnvironmentalResearch2017], which substantially overlaps but is not identical to the @ElliottValuesEnvironmentalResearch2017 scale.  On the METI, participants rate the target scientist (the fictional Dr. Spence) on a 1-7 scale for 14 items, where each item is anchored at the ends by a pair of words, such as competent-incompetent or responsible-irresponsible.  In both @ElliottValuesEnvironmentalResearch2017 and our replication analysis, all items were averaged together to form a 1-7 composite measure of trustworthiness.  To aid interpretation, in the analysis the direction of the scale has been set so that increased values correspond to increasing perceived trustworthiness.

Ethicists make a key distinction between trust and trustworthiness *[Baier]*.  Trust is a verb: placing one's trust in someone else, with respect to some activity or domain.  Trustworthiness is an assessment, of whether or not that trust is appropriate.  In the experiment, participants might trust Dr. Spence's conclusion even if they judge Spence to be untrustworthy, or vice versa; though we would typically expect these two to go together.  As its name indicates, METI assesses perceived trustworthiness, not trust (such as accepting Dr. Spencer's conclusion about BPA).

@ElliottValuesEnvironmentalResearch2017 explain that they selected BPA as a complex, ongoing, public scientific controversy.  For our replication, we chose to keep BPA, rather than switching to a different public scientific controversy with a higher profile in 2021, such as climate change, police violence, voter fraud, or any of numerous aspects of the Covid-19 pandemic.  All of these controversies are highly politically charged, with prominent public experts and counterexperts *[cite]*.  While we anticipated some effects of political partisanship in the BPA case, we felt it would be less likely to swamp the values disclosure that was our primary interest.

@@

Participants were recruited using the online survey platform Prolific, and the survey was administered in a web browser using Qualtrics.  Prolific has an option to draw samples that are balanced to be representative by age, binary gender, and a 5-category race variable (taking values Asian, Black, Mixed, Other, and White) for US adults *[https://researcher-help.prolific.co/hc/en-gb/articles/360019238413-Representative-samples-FAQ]*.  A recent analysis finds that Prolific produces substantially higher quality data than Amazon Mechanical Turk for online survey studies, though three of the five authors are affiliated with Prolific [@PeerDataQualityPlatforms2021].  Preliminary power analysis recommended a sample of approximately 1,000 participants to reliably detect non-interaction effects (hypotheses a-d).

@@

*[All original data and reproducible code is available at <https://github.com/dhicks/transparency>.  Instructions in that repository explain how to automatically reproduce our analysis.]*

@@

Critically, our data are unlikely to be representative by education level and political ideology.  In 2021, about 9% of US adults 25 or over had a less than high school education, and 38% had a Bachelor's degree or higher.  Only 1% of our participants reported a less than high school education, and 57% reported a Bachelor's degree or higher.  *[<https://www.census.gov/library/visualizations/time-series/demo/cps-historical-time-series.html> fig. 2]*  For political ideology, the General Social Survey has consistently found over several decades that about 30% of US adults identify as liberal, about 30% identify as conservative, and about 40% as moderate *[https://gssdataexplorer.norc.org/variables/178/vshow]*.  Among our participants, liberals outnumber conservatives *[by over 2:1; hypothesis a fig]*.  Both overrepresentation of college graduates and underrepresentation of conservatives (especially conservatives with strong anti-institutional views) are known issues in public opinion polling *[https://academic.oup.com/poq/article/82/1/1/4837043?login=false]*.  In exploratory data analysis, we noted that there was essentially no correlation between political ideology and perceived trustworthiness *[supplement]*.  This was surprising, since general trust/distrust in science has become a partisan phenomenon over the last few decades:  General Social Survey data shows increasing trust in science from liberals and decreasing trust from conservatives [@GauchatPoliticizationSciencePublic2012; @LeePartyPolarizationTrust2021], and many (though not all) prominent public controversies align with liberal-conservative partisanship *[cite]*.  Insofar as some political conservatives are both less likely to participate in studies on Prolific (or at least in our particular study) and likely to perceive Spence as less trustworthy, this will lead to omitted variable bias for analyses that require adjustment by political ideology.  *[however, most of our analyses do not require this adjustment]*

@@

We ran the following analyses to test our hypothesis that (H5) the magnitude of the effects found for the tests of H2-H4 vary depending on whether the participant prioritizes public health or economic growth. Results are presented in Table #.

@@

>>>>

---
title: "Something something values disclosure replication"
author: "Daniel J. Hicks and Emilio Lobato"
abstract: |
  abstract
toc: true
execute:
    echo: false
bibliography: |
  transparency-Hicks.yaml
---

@@

Over the past 15 years, many philosophers of science have rejected the ideal of value-free science and related, traditional understandings of objectivity and political neutrality of science [@DouglasSciencePolicyValuefree2009; @ElliottTapestryValuesIntroduction2017].  According to the value-free ideal, social and political values — such as feminism, environmentalism, or the protection of human health — have no legitimate role to play in the evaluation of scientific hypotheses.  The value-free ideal is compatible with allowing social and political values to play important roles earlier and later in inquiry.  Specifically, these values may legitimately shape the content and framing of research questions — researchers might decide to investigate whether chemical X causes cancer out of a concern to protect human health — and will be essential when scientific findings are used to inform public policy — say, banning the use of chemical X.  But, according to the value-free ideal, these values must not influence the collection and analysis of data and the process of reaching an overall conclusion about whether or not chemical X causes cancer.

Challenges to the value-free ideal argue that at least some social and political values may, or even should, play a role in the evaluation of scientific hypotheses.  Keeping with the example, the question of whether chemical X causes cancer has much more significant social and political implications than the question of whether chemical X fluoresces under ultraviolet light.  So, in the interest of protecting human health, it would be appropriate to *scientifically* accept the hypothesis that chemical X is carcinogenic based on relatively provisional evidence [@CranorSocialBenefitsExpedited1995; @ElliottNonepistemicValuesMultiple2014; @FernandezPintoLegitimizingValuesRegulatory2019].

While this kind of argument indicates that at least some social and political values may or should play at least some role in the evaluation of at least some scientific hypotheses, it doesn't provide us with much positive guidance:  which values, playing which roles, in the evaluation of which hypotheses?  [@HicksNewDirectionScience2014]  One (partial) answer to this set of questions appeals to the ideal of transparency:  scientists should disclose to the public the social and political values that have influenced their research [@ElliottSciencePolicyTransparency2014; @McKaughanBacktrackingEthicsFraming2013].  But some philosophers have objected to this transparency proposal, arguing that it might actually undermine trust in science.  According to this objection, members of the public generally accept the ideal of value-free science; values disclosures would violate this ideal, making scientists (incorrectly) appear biased and untrustworthy [@JohnEpistemicTrustEthics2017; @KovakaClimateChangeDenial2021].

@@

| No other information about the (fictional) Dr. Spence is provided.  The content of the slide varies across the conditions.  Each slide has the header "My conclusion," followed by a list of 2 or 3 bullets.  The first bullet, if present, makes a values disclosure, stating that either economic growth or public health should be a top national priority.  This first bullet is not present in the "no disclosure" values condition.  The second bullet, identical across conditions, states "I examined the scientific evidence on potential health risks of BPA."  The third bullet concludes that BPA either does or does not cause harm to people, depending on whether the subject is in the "causes harm" or "does not cause harm" condition.

After viewing the slide, the participant is then asked to rate Dr. Spence's trustworthiness using a semantic differential scale.  @ElliottValuesEnvironmentalResearch2017 did not cite a source for their scale; we used the Muenster Epistemic Trustworthiness Inventory [METI\; @HendriksMeasuringLaypeopleTrust2015], which substantially overlaps but is not identical to the @ElliottValuesEnvironmentalResearch2017 scale.  On the METI, participants rate the target scientist (the fictional Dr. Spence) on a 1-7 scale for 14 items, where each item is anchored at the ends by a pair of words, such as competent-incompetent or responsible-irresponsible.  In both @ElliottValuesEnvironmentalResearch2017 and our replication analysis, all items were averaged together to form a 1-7 composite measure of trustworthiness.  To aid interpretation, in the analysis the direction of the scale has been set so that increased values correspond to increased perceived trustworthiness.

Ethicists make a key distinction between trust and trustworthiness [@BaierTrustAntitrust1986].  Trust is a verb: placing one's trust in someone else, with respect to some activity or domain.  Trustworthiness is an assessment, of whether or not that trust is appropriate.  In the experiment, participants might trust Dr. Spence's conclusion even if they judge Spence to be untrustworthy, or vice versa; though we would typically expect these two to go together.  As its name indicates, METI assesses perceived trustworthiness, not trust (such as accepting Dr. Spencer's conclusion about BPA).

@ElliottValuesEnvironmentalResearch2017 explain that they selected BPA as a complex, ongoing, public scientific controversy.  For our replication, we chose to keep BPA, rather than switching to a different public scientific controversy with a higher profile in 2021, such as climate change, police violence, voter fraud, or any of numerous aspects of the Covid-19 pandemic.  All of these controversies are highly politically charged, with prominent public experts and counterexperts [@GoldenbergVaccineHesitancyPublic2021 100-1, ch. 6].  While we anticipated some effects of political partisanship in the BPA case, we felt it would be less likely to swamp the values disclosure that was our primary interest.

@@

Participants were recruited using the online survey platform Prolific, and the survey was administered in a web browser using Qualtrics.  Prolific has an option to draw samples that are balanced to be representative by age, binary gender, and a 5-category race variable (taking values Asian, Black, Mixed, Other, and White) for US adults *[https://researcher-help.prolific.co/hc/en-gb/articles/360019238413-Representative-samples-FAQ]*.  A recent analysis finds that Prolific produces substantially higher quality data than Amazon Mechanical Turk for online survey studies, though three of the five authors are affiliated with Prolific [@PeerDataQualityPlatforms2021].  Preliminary power analysis recommended a sample of approximately 1,000 participants to reliably detect non-interaction effects (H1-4).

@@

*[Anonymized original data and reproducible code are available at <https://github.com/dhicks/transparency>.  Instructions in that repository explain how to automatically reproduce our analysis.]*

@@

Critically, our data are unlikely to be representative by education level and political ideology.  In 2021, about 9% of US adults 25 or over had a less than high school education, and 38% had a Bachelor's degree or higher.  Only 1% of our participants reported a less than high school education, and 57% reported a Bachelor's degree or higher.  *[<https://www.census.gov/library/visualizations/time-series/demo/cps-historical-time-series.html> fig. 2]*  For political ideology, the General Social Survey has consistently found over several decades that about 30% of US adults identify as liberal, about 30% identify as conservative, and about 40% as moderate *[https://gssdataexplorer.norc.org/variables/178/vshow]*.  Among our participants, liberals outnumber conservatives *[by over 2:1]* (@fig-part-values).  Both overrepresentation of college graduates and underrepresentation of conservatives (especially conservatives with strong anti-institutional views) are known issues in public opinion polling [@KennedyEvaluation2016Election2018].  In exploratory data analysis, we noted that there was essentially no correlation between political ideology and perceived trustworthiness *[supplement]*.  This was surprising, since general trust/distrust in science has become a partisan phenomenon over the last few decades:  data from the General Social Survey shows increasing trust in science from liberals and decreasing trust from conservatives [@GauchatPoliticizationSciencePublic2012; @LeePartyPolarizationTrust2021], and many (though not all) prominent public controversies align with liberal-conservative partisanship *[cite]*.

Because of these representation issues, insofar as some political conservatives are both less likely to participate in studies on Prolific (or at least in our particular study) and likely to perceive Spence as less trustworthy, this will produce omitted variable bias for analyses that require adjustment by political ideology.  Analysis of our a priori DAG indicated that, for the hypotheses examined here, this adjustment was not necessary in any case.  When some adjustment was necessary (H4-5), we included political ideology along with other demographic variables in an alternative model specification as a robustness check.  In line with the DAG, in no case did these robustness checks produce indications of omitted variable bias.

@@

We ran the following analyses to test our hypothesis that (H5) the magnitude of the effects found for the tests of H2-H4 vary depending on whether the participant prioritizes public health or economic growth. Results are presented in [@tbl-eb; @tbl-ec; @tbl-sci-part].

@@

@tbl-summary summarizes the results of our replication attempts. We were able to successfully replicate the correlation between political ideology and valuing public health and consumer risk sensitivity (a scientist who concludes BPA is harmful is perceived as more trustworthy), and in addition found evidence of an interaction between consumer risk sensitivity and whether the participant values public health (participants who value public health have more extreme reactions to the scientist's conclusion).

We did not find evidence for a transparency penalty or shared values effect.  And we found evidence of a scientist values effect, which also did not appear in @ElliottValuesEnvironmentalResearch2017.

+------------------+------------------------------------------+-------------+
|                  | Hypothesis                               | Replicated? |
+==================+==========================================+=============+
|              H1a | Liberals prioritize public health        | yes         |
|                  | rel. to conservatives                    |             |
+------------------+------------------------------------------+-------------+
|              H1b | Majority of conservatives prioritize     | no          |
|                  | public health                            |             |
+------------------+------------------------------------------+-------------+
|               H2 | Sci. who find chemical is harmful        | yes         |
|                  | perceived as more trustworthy            |             |
+------------------+------------------------------------------+-------------+
|               H3 | Sci. who disclose values perceived as    | no          |
|                  | less trustworthy                         |             |
+------------------+------------------------------------------+-------------+
|               H4 | Given sci. discloses values,             | no          |
|                  | sci. who shared values with participant  |             |
|                  | perceived as more trustworthy            |             |
+------------------+------------------------------------------+-------------+
|        Unplanned | Given sci. discloses values,             |yes$^\dagger$|
|                  | sci. who values public health            |             |
|                  | perceived as more trustworthy            |             |
+------------------+------------------------------------------+-------------+
|      H5-consumer | Interaction btwn. sci. conclusion (H2)   | yes         |
|                  | and participant values                   |             |
+------------------+------------------------------------------+-------------+
|  H5-transparency | Interaction btwn. values disclosure (H3) | no          |
|                  | and participant values                   |             |
+------------------+------------------------------------------+-------------+
|        H5-shared | Interaction btwn. shared values (H4)     | $^\star$    |
|                  | and participant values                   |             |
+------------------+------------------------------------------+-------------+
|   H5-sci. values | Interaction btwn. scientist values       | no$^\dagger$|
|                  | and participant values                   |             |
+------------------+------------------------------------------+-------------+

: Summary of replication results.  $^\star$: The shared values $\times$ participant values interaction specification had perfect multicollinearity and could not be fit.  $^\dagger$: For these unplanned analyses, "yes" indicates that we found evidence supporting the stated hypothesis, and "no" indicates that we did not find such evidence. {#tbl-summary}

@ElliottValuesEnvironmentalResearch2017 claim evidence for a shared values effect by comparing values disclosure vs. no disclosure (that is, estimating a transparency effect) in quasi-independent subsamples across 8 cells (participant values $\times$ scientist conclusion $\times$ scientist values; @ElliottValuesEnvironmentalResearch2017 fig. 1).  This approach is difficult to interpret; their reasoning seems to be that this estimate is statistically significant in 4/8 cells, and in 3/4 of these cells (1 where respondent values economic growth) the participant and scientist share values.  But these analyses are underpowered (for example, sample sizes are well below 100 for some of the cells in which participants value economic growth) and the overall approach cannot distinguish a shared values effect from other potential effects (scientist values, participant values).  Applying our regression analysis approach to the data published by @ElliottValuesEnvironmentalResearch2017, the estimate for shared values was not statistically significant, $\beta$=0.29, t(335)=1.75, p=0.08; while the estimate for scientist values was statistically significant, $\beta$=0.41, t(335)=2.48, p=0.014.  It seems likely that claims of a shared values effect were the result of misinterpreting a confusing statistical approach.  *[trying to put this as nice as possible]*

The disagreement over a transparency penalty is more difficult to explain.  Using our approach and the data from @ElliottValuesEnvironmentalResearch2017, the estimate for a values disclosure was statistically significant (@tbl-bc model 1).  One possible explanation is that, over the course of the Covid-19 pandemic, members of the general public have become used to "scientists" (including both bench researchers and public health officials) making claims about the importance of protecting public health.  (And, more recently, Leana Wen and some other prominent physicians and public health officials have argued for the importance of a "return to normal," which might be understood as prioritizing economic growth.)  A values disclosure that was regarded as violating the value-free ideal, pre-pandemic, might now be seen as routine.

*[further work]*

# References #
